---
title: "class09"
author: "Mason Lew (PID: A17533139)"
format: gfm
toc: TRUE
---

Today we will examine data from 538 on common halloween candy. In particular we will use ggplot, 

```{r}
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"

candy <- read.csv(url, row.names = 1)

head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
ncol(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

> How many chocolate candies are there

```{r}
sum(candy$chocolate)
```

>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["M&M's", "winpercent"]
```

```{r}
candy["M&M's",]$winpercent
```

>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

**N.B.** The `winpercent` variable is on a different scale (about a 100x magnitude/ 1-100% versus 0-1). We will most likely need to scale the data if we plan to do PCA.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

the 0s are not chocolate while the 1s are chocolate

>Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent) +
  geom_histogram(bins = 10, fill = "navy")

```


>Q9. Is the distribution of winpercent values symmetrical?

No

>Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```

mean is above 50 but median is below 50 at 47.38. The center of the distribution is more so the median so, no the center is not above 50.

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

-step one: find all "chocolate" candy
-step two: find their "winpercent" values
-step three: summarize these values
-step four: find all "fruit" candy
-step five: find their "winpercent" values
-step six: summarize these values
-step seven: compare summarized points

1. Find all chocolate candy
```{r}
choc.inds <- candy$chocolate == 1
```

2. Find their "winpercent" values
```{r}
choc.win <- candy[choc.inds,]$winpercent
```

3. Summarize these values
```{r}
choc.mean <- mean(choc.win)
choc.median <- median(choc.win)

choc.mean
choc.median
```
4. Find all fruit candy
```{r}
fruit.inds <- candy$fruity == 1
```

5. Find their "winpercent"
```{r}
fruit.win <- candy[fruit.inds,]$winpercent
```

6. Summarize these values
```{r}
fruit.mean <- mean(fruit.win)
fruit.median <- median(fruit.win)

fruit.mean
fruit.median
```

7. Compare
```{r}
choc.mean
fruit.mean

choc.median
fruit.median
```

On average, across both mean and median, Chocolate had a higher `winpercent`

>Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruit.win)
```

Yes! The difference is statistically significant based on a ttest at 95% confidence resulting in a p-value <.05.

>Q13. What are the five least liked candy types in this set?

```{r}
#not that useful - it just sorts values
sort(candy$winpercent)
```

```{r}
x <- c(10, 1, 100)
order(x)
```

```{r}
x [order(x)]
```

The `order()` function tells us how to arrange the elements of the input to make them sorted - i.e. how to order them

We can determine the order of winpercent to make them sorted and use that order to arrange the whole data set

```{r}
ord.inds <- order(candy$winpercent)
head(candy[ord.inds,])
```


>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
ord.inds <- order(candy$winpercent)
tail(candy[ord.inds,])
```

```{r}
ord.inds.d <- order(candy$winpercent, decreasing = TRUE)
head(candy[ord.inds.d,])
```

> Q15. Make a first barplot of candy ranking based on winpercent values.


```{r}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```

We need to make our own separate color vector where we can spell out exactly what candy is colored a particular color

```{r}
mycols <- rep("black", nrow(candy))
mycols[candy$chocolate == 1] <- "chocolate"
mycols[candy$bar == 1] <- "brown"
mycols[candy$fruity == 1] <- "lightblue"
```

```{r}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col(fill = mycols)
```

> Q17. What is the worst ranked chocolate candy?

Hershey's Milk Chocolate

> Q18. What is the best ranked fruity candy?

Airheads

## Taking a look at price percent

Make a plot of winpercent (x-axis) vs pricepercent (y-axis)
```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label = rownames(candy)) +
  geom_point(col = mycols) +
  geom_text()
```

to avoid over plotting of text we can use the `repel` package.

```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label = rownames(candy)) +
  geom_point(col = mycols) +
  geom_text_repel(max.overlaps = 12) +
  theme_bw()
```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

The least popular of the top 5 most expensive candies is Nik L Nip

> Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```


## 5. Exploring the correlation structure

Now that we have explored the dataset a little, we will see how the variables interact with one another.

First we will use correlation and view the results with the **corrplot** package to plot a correlation matrix
```{r}
cij <- cor(candy)
```

```{r}
library (corrplot)

corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

fruit and chocolate are anti-correlated (but any red values)

> Q23. Similarly, what two variables are most positively correlated?

variables are correlated with themselves. The variables that are positively correlated would be chocolate and winpercent and chocolate with pricepercent (anything in blue)

##PCA

Let's applyy PCA using `prcomp` 
```{r}
pca <- prcomp(candy, scale = TRUE)
```

```{r}
summary(pca)
```

```{r}
attributes(pca)
```

Let's plot our main results as our PCA "score plot"

```{r}
p <- ggplot(pca$x) +
  aes(PC1, PC2, label = rownames(pca$x)) +
  geom_point(col=mycols) 
p

```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=mycols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

Finally lets look at how the original variables contribute to the PCs, Start with PC 1

```{r}
ggplot(pca$rotation) + 
  aes(PC1, rownames(pca$rotation)) +
  geom_col()
```


> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, and pluribus. Yes, this makes sense to me as these match the results of the correlation plot from earlier.
